---
title: "Udacity Project: Explore and Summarize Data"
author: "Matthew Dolder"
date: "5/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1 Analysis and Exploration.

    a. Headings and text should organize your thoughts and reflect your analysis as you explored the data.

    b. Plots in this analysis do not need to be polished with labels, units, and titles; these plots are exploratory (quick and dirty). They should, however, be of the appropriate type and effectively convey the information you glean from them.

    c. You can iterate on a plot in the same R chunk, but you don’t need to show every plot iteration in your analysis.

# Section 2 Final Plots and Summary

    You will select three plots from your analysis to polish and share in this section. The three plots should show different trends and should be polished with appropriate labels, units, and titles (see the Project Rubric for more information).

    A final section called “Reflection”

    This should contain a few sentences about your struggles, successes, and ideas for future exploration on the data set (see the Project Rubric for more information).

# Section 3 - Reflection
This should contain a few sentences about your struggles, successes, and ideas for future exploration on the data set (see the Project Rubric for more information).

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
