Prosper Loan Analysis by Matthew Dolder
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.



library(ggplot2)
library(dplyr)
library(sqldf)
library(lubridate)
library(scales)
library(gridExtra)
```



```{r echo=FALSE, Load_the_Data,  warning=FALSE}
# Load the Data
loans<-read.csv("prosperLoanData.csv")

```


Reference: 
https://www.nerdwallet.com/reviews/loans/personal-loans/prosper-personal-loans

Prosper is peer-to-peer lending institution.  Borrowers apply for loans much 
like a bank loan and any number of people can invest in the loan expecting 
a return.  

Looking at the csv file, I noticed a wide range in the number of investors.  

 Investors      
 Min.   :   1.00  
 1st Qu.:   2.00  
 Median :  44.00  
 Mean   :  80.48  
 3rd Qu.: 115.00  
 Max.   :1189.00  

Why do some loans attract so many investors while many, over 27,800 attract
only 1 investor?  Are investors looking for high risk loans which pay a 
higher APR?  Are they looking for low risk loans?  Are the motivations 
even financial?  Perhaps they want to support teachers, nurses, and fireman to 
feel they are doing a service to the community.  Perhaps they favor homeowners.  
Perhaps it's simply the length of time since the 
loan was created, giving more time for investors to jump on board.  

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

#examine size of dataset
tcount<-nrow(loans) 
tcount
#There are 113,937 loans in the dataset.

#The number of variables is massive
#Here I limit the dataset to the variables we are interested in
#That will make it easier to investigate the data and perhaps improve
#performance
loanvars<-c("Investors","Occupation","DebtToIncomeRatio","ProsperScore",
"LoanOriginalAmount","Term","LoanStatus","ListingCreationDate",
"LoanMonthsSinceOrigination","TotalProsperLoans","BorrowerAPR",
"CreditScoreRangeUpper","IsBorrowerHomeowner","EmploymentStatus")
loans<-loans[loanvars]
summary(loans)


```




```{r echo=FALSE, message=FALSE, Univariate_Plot_1, warning=FALSE}


#show histogram of investor count
ggplot(data=subset(loans,Investors < 400),aes(x=Investors)) + 
  geom_histogram(color = I('black'), fill = I('#099DD9'))
#most loans have 1 investor

```

Figure 1. Histogram of the Investor variable.

This graph is capped at 400 investors.  The max number of investors for a single
loan is 1,189.  The graph shows that most loans have 1 investor only.  


```{r echo=FALSE, message=FALSE, Univariate_Plot_2, warning=TRUE}
#reference: 
#https://discuss.analyticsvidhya.com/t/how-to-count-number-of-distinct-values-in-a-column-of-a-data-table-in-r/1124/2
#sqldf("select distinct(Occupation) from loans")  
#68 distinct Occupations returned

occupation <- loans %>%
  group_by(Occupation) %>%
  summarize(loan_count = n()) %>%
  arrange(desc(loan_count)) %>%
  slice_head(n=20)

#show count of each occupation
#x = reorder(day, -perc)
  ggplot(data=occupation,aes(y=reorder(Occupation,loan_count), x=loan_count)) + 
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9'),position="stack") +
  ylab("occupation") + xlab("loan count")

empstatus <- subset(loans,Occupation="Other") %>%
  group_by(EmploymentStatus) %>%
  summarize(loan_count = n()) %>%
  arrange(desc(loan_count)) 

 
  ggplot(data=empstatus,aes(y=reorder(EmploymentStatus,loan_count), x=loan_count)) + 
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9'),position="stack") +
  ylab("employment status for (Other)") + xlab("loan count")


```

Figure 2. Number of Loans by Occupation. 
Figure 3. Number of Loans by employment status where Occupation = "Other"

Figure 2 shows that "Other" is the most popular occupation.  This isn't
very helpful for analysis.  Figure 3 shows that for loans where occupation
is other, we can tease more information from the employment status such as 
"self-employed" or "retired".  


```{r echo=FALSE, message=FALSE, Univariate_Plot_4, warning=FALSE}

#show histogram of DebtToIncomeRation
ggplot(data=subset(loans,DebtToIncomeRatio < 1),aes(x=DebtToIncomeRatio)) + 
  geom_histogram(color = I('black'), fill = I('#099DD9'))  

```

Figure 4. Loan count by Debt to Income Ration

This graph shows that Prosper has a preference towards borrowers with a 
Debt to Income ratio between 0.1 and 0.3. 


```{r echo=FALSE, message=FALSE, Univariate_Plot_5, warning=FALSE}

#show ProsperScore
ggplot(data=loans,aes(x=factor(ProsperScore))) + 
  geom_bar(color = I('black'), fill = I('#099DD9')) +
  xlab("Prosper Score")
#only available for 2009 and later

```

Figure 5. Loan count by Prosper score. 

Prosper score is the result of risk formula applied to loans starting in 2009.
Loans originating prior to 2009 show as "NA". 
  

```{r echo=FALSE, message=FALSE, Univariate_Plot_6, warning=FALSE}

#Show histogram of loan size
ggplot(data=loans,aes(x=LoanOriginalAmount)) + 
  geom_histogram(color = I('black'), fill = I('#099DD9'))  

```

Figure 6.  Loan count by Amount in Dollars.  

Few loans are above $25,000.  


```{r echo=FALSE, message=FALSE, Univariate_Plot_7, warning=FALSE}

#Show loan terms and number of each.
ggplot(data=loans,aes(x=factor(Term))) + 
  geom_bar(color = I('black'), fill = I('#099DD9')) + 
  xlab("Term")

```

Figure 7.  Loan count by Term (length of loan)

36 months is the most popular term.  


```{r echo=FALSE, message=FALSE, Univariate_Plot_8, warning=FALSE}

#show number of loans with each status
ggplot(data=loans,aes(y=LoanStatus)) + 
  geom_bar(color = I('black'), fill = I('#099DD9'))  

```

Figure 8.  Loan count by Status.  

Current and Complete are the most popular categories.  

```{r echo=FALSE, message=FALSE, Univariate_Plot_9, warning=FALSE}
#parse year from loan start date and put in a new variable
loans <- loans %>% mutate(ListingCreationYear = year(ListingCreationDate))
#This will make grouping by year easier later. 

#reference
#https://stackoverflow.com/questions/50935857/ggplot2-keeps-adding-5-at-end-of-year-variable-on-x-axis
#show number of loans per year. 
ggplot(data=loans,aes(x=ListingCreationYear)) + 
  geom_bar(color = I('black'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = 0:2100)


```


Figure 9.  Loans issued by year.  

The dataset begins in November 2005, hence the small number there.  
In 2009 changes in banking regulations due to the recession clearly had an affect.  
Prosper has grown considerably since.  The data set ends in March 2014.


```{r echo=FALSE, message=FALSE, Univariate_Plot_10, warning=FALSE}

#show current length of loan
ggplot(data=subset(loans,loanstatus="Current"),aes(x=LoanMonthsSinceOrigination)) + 
  geom_bar(color = I('black'), fill = I('#099DD9')) 
#The largest term length is 60 months.
#However Prosper continues counting months after the loans are completed. 

#Add a variable called "months_active" which stops at the end of the term. 
loans <- loans %>% 
  mutate(months_active = 
  ifelse(LoanMonthsSinceOrigination > Term,Term,LoanMonthsSinceOrigination))

```

Figure 10.  Number of loans by months since origination. 

This is a strange variable because the months keep counting up after the loan 
is complete.  


```{r echo=FALSE, message=FALSE, Univariate_Plot_11, warning=FALSE}

#Number of Prosper loans the borrower at the time they created this listing. 
#This value will be null if the borrower had no prior loans. 
ggplot(data=subset(loans,loanstatus="Current"),aes(x=TotalProsperLoans)) + 
  geom_bar(color = I('black'), fill = I('#099DD9')) 

```

Figure 11.  Number of loans by 'total prosper loans'

The TotalProsperLoans variable is the number of Prosper loans the borrower 
previously had.  My assumption is that preference is given to repeat customers.  
The majority of loans are first time borrowers.  

```{r echo=FALSE, message=FALSE, Univariate_Plot_12, warning=FALSE}

#The Borrower's Annual Percentage Rate (APR) for the loan.
ggplot(data=loans,aes(x=BorrowerAPR)) + 
  geom_histogram(color = I('black'), fill = I('#099DD9')) 

```

Figure 12. Number of loans by APR

Generally, a higher APR is given to loans with a higher risk and can also leads 
to a higher return.   

```{r echo=FALSE, message=FALSE, Univariate_Plot_13, warning=FALSE}

#Plot number of loans with max DTI as defined by the loan documentation
ggplot(data=subset(loans,loans$DebtToIncomeRatio==10.01),
       aes(x=Investors)) +
  geom_histogram(color = I('black'), fill = I('#099DD9'))


```

Figure 13.  Number of investors where debt to income ratio is 10.01

I found it interesting that anyone would invest in a loan with such a 
high debt to income ratio.  

# Univariate Analysis


### What is the structure of your dataset?

The dataset contains 80 variables and nearly 114,000 observations.  
The variables describe the terms of the loan, the repayment status, 
and the financial health of the borrower.  


### What is/are the main feature(s) of interest in your dataset?

I'm interested in information about the borrower or the loan which 
influences investors to choose a loan.


### What other features in the dataset do you think will help support your 

The data spans the 2008 recession. Comparing loans before and after may 
show a shift in investor's behavior or even a change lending regulations. 
Prosper Score was added in 2009 as a result of the recession and new 
regulations.  Does this impact the number of investors? 


### Did you create any new variables from existing variables in the dataset?

ListingCreationYear: Contains year parsed from ListingCreationDate
months_active:  Number of months since loan was created, but capped at the 
                term of the loan.  
                
# Bivariate Plots Section

What attracts investors to loans?  Are they choosing wisely?  Loan 
Status vs Investor count could show us if they are choosing wisely.  
Do loans in default have a high number of investors?  

Debt to income ratio (DTI) looks interesting.  The population of total loans are
grouped around a seemingly low DTI with a few at the high end.  In the 
bivariate plots section, I want to compare investors to DTI and see 
if the number of investors has a similar graph as the total number of loans 
or if investors are disproportionately flocking toward borrowers with a higher 
risk.  

I'm also interested in investors per year because the data spans a period 
both before and after the 2008 recession.  Were investors careless before 2008
and more conservative after?  

Does occupation impact the number of investors? Perhaps they aren't choosing
based on calculations, but instead feel a need to support teachers. 

I'm not as interested in Prosper score as there is no data prior to 2009.



```{r echo=FALSE, Bivariate_Plots_1, warning=FALSE}

###################################################
#using dplyr, summarize Investors by loan status 
#reference:  Udacity Lesson 6: Explore two variables
loans.investors_by_status <- loans %>%
  group_by(LoanStatus) %>%
  summarize(mean_investors = mean(Investors),
            median_investors = median(Investors),
            n=n())
#Plot the mean number of investors by loan status from the previous statement.
p1 <- ggplot(data=loans.investors_by_status,aes(x=LoanStatus,y=n)) + 
  geom_col(color = I('black'), fill = I('#099DD9')) + 
  scale_x_discrete(guide = guide_axis(n.dodge=3)) + 
  ylab("number of investors")

p2 <- ggplot(data=loans.investors_by_status,aes(x=LoanStatus,y=mean_investors)) + 
  geom_col(color = I('black'), fill = I('#099DD9')) + 
  scale_x_discrete(guide = guide_axis(n.dodge=3)) + 
  ylab("mean number of investors")

grid.arrange(p1,p2)
####################################################

```

Figure 14. Investors by Loan Status

This graph shows the total number of investors by loan status.  Status of 
"Current" has the most number of loans so the top graph is not surprising.  
Looking at the mean investors per status, we can see loans which performed 
poorly can have many investors per loan.  

```{r echo=FALSE, Bivariate_Plots_2, warning=FALSE}

####################################################
#summarize Investors by Occupation
loans.investors_by_occupation <- loans %>%
  group_by(Occupation) %>%
  summarize(mean_investors = mean(Investors),
            median_investors = median(Investors),
            n=n()) %>%
  arrange(mean_investors)

#since there are so many occupations, focus on those above the 
#3rd quantile in mean_investors
summary(loans.investors_by_occupation$mean_investors) #93.66

#subset previous grouping by 3rd quartile
occupation_with_most_investors <- 
  subset(loans.investors_by_occupation,mean_investors > 93.66)

#plot the mean number of investors by occupation for 3rd quartile
ggplot(data=occupation_with_most_investors,
       aes(x=mean_investors,y=reorder(Occupation,mean_investors,sum))) +
  geom_col(color = I('black'), fill = I('#099DD9')) +
  ylab("occupation") + 
  xlab("mean investors per occupation")

#########################################################

```

Figure 15. Mean number of investors per occupation

This graph shows the occupation of borrowers which investors are most 
interested in.  Looking at this alone, we can't tell if investors are 
choosing by occupation or if these occupations just happen to have the 
best credit score.  

```{r echo=FALSE, Bivariate_Plots_3, warning=FALSE}

#########################################################
#summarize Investors by credit score
loans.credit_score_by_occupation <- loans %>%
  group_by(Occupation) %>%
  summarize(mean_credit = mean(CreditScoreRangeUpper),
            median_credit = median(CreditScoreRangeUpper),
            n=n()) %>%
  arrange(mean_credit)

#grab the 3rd quartile
summary(loans.credit_score_by_occupation$mean_credit)
#mean_credit 3rd Qu.:717.7

#subset by 3rd quertile
occupation_with_highest_credit_score <- 
  subset(loans.credit_score_by_occupation,mean_credit > 717.7)

#Plot the mean credit score by occupation  
ggplot(data=occupation_with_highest_credit_score,
       aes(x=mean_credit,y=reorder(Occupation,mean_credit,sum))) +
  geom_col(color = I('black'), fill = I('#099DD9')) +
  ylab("occupation") + 
  xlab("mean credit score by occupation")
#Examine overlap with the previous plot.  Do the same occupations
#appear at the top? 
#########################################################

```

Figure 16. Mean credit score by occupation

This graph shows the 18 occupations where borrowers have the highest 
mean credit score.  The different in credit score between each one is trivial. 
What is interesting is the overlap of occupations with Figure 15.  
With a few exceptions, such as Fireman and Military Officer, the occupations 
with the highest mean number of investors also have the highest mean credit score.  


```{r echo=FALSE, Bivariate_Plots_4, warning=FALSE}


#########################################################
#Plot the number of Investors by credit score
ggplot(data=subset(loans,loans$CreditScoreRangeUpper>400 
                   & loans$Investors < 1000 & 
                     loans$Investors > 0),
       aes(x=Investors,y=CreditScoreRangeUpper)) +
  geom_point(alpha=1/100) +
  ylab("credit score range - upper")
#There appears to be a cutoff around credit score 700.
#the number of investors starts dropping.  There is also 
#a cutoff at the very top range.  
########################################################

```

Figure 17.  Number of investors by credit score where credit score is above 
            400 and number of investors is greater than 0 and less than 1000.

This graph shows a high concentration of loans and investors around the 750
credit score mark.  Few loans have a very high number of investors and 
few have a credit score below 500.  

I would expect this graph to be the most dense at the top of the credit score 
range.   My guess is that few people have a perfect score and very few of 
those people are interested in borrowing money from Prosper.  



```{r echo=FALSE, Bivariate_Plots_5, warning=FALSE}

########################################################
#Plot investors by months_active
ggplot(data=loans,
       aes(x=Investors,y=months_active)) +
  geom_point(alpha=1/100) +
  ylab("months active")
#There appears to be a huge spike at 36months
########################################################

```

Figure 18.  Number of investors by months active.  

Months active is a custom variable which shows takes the number of months
since loan origination and caps it at the term of the loan.  I wanted to 
see if more investors jump on board a loan as time goes on.  It 
appears that is not the case, however, there is a higher concentration 
of investors at 12 & 36 months which are popular term lengths.  

```{r echo=FALSE, Bivariate_Plots_6, warning=FALSE}

########################################################
#Plot investors by Debt to Income Ratio
#I'll abbreviate this as DTI in the future
ggplot(data=loans,
       aes(x=Investors,y=DebtToIncomeRatio)) +
  geom_point(alpha=1/20) +
  ylab("debt to income ratio")

```

Figure 19.  Number of investors by debt to income ratio. 

This graph shows that out of the entire population of loans, investment
is heavy below 0.1 debt to income.  It also shows an unusual outlier at the top 
end of debt to income ratio.  These are loans where debt to income is greater
than 100%.  Why would anyone invest in such risky loans?  Perhaps 
there is another factor such as occupation.  

```{r echo=FALSE, Bivariate_Plots_7, warning=FALSE}

#plot Investors by occupation for max DTI
ggplot(data=subset(loans,loans$DebtToIncomeRatio==10.01 & loans$Investors>100),
       aes(x=Investors,reorder(Occupation,Investors,sum))) +
  geom_col() + 
  ylab("occupation")
#by far Other is the most popular Investment category in 100% DTI

```

Figure 20.  Number of investors by occupation where debt to income ratio = 10.01

Figure 20 expands on Figure 19 by looking at borrowers with a maximum 
debt to income ratio.  Unfortunately, the largest category by far is "other". 
This doesn't tell us much about the borrower.  

```{r echo=FALSE, Bivariate_Plots_8, warning=FALSE}

#plot Investors by employment status for Max DTI 
ggplot(data=subset(loans,
                   loans$DebtToIncomeRatio==10.01 & loans$Occupation=="Other"),
       aes(x=Investors,y=EmploymentStatus)) +
  geom_col() + 
  ylab("employment status")

#This helps to break up the Other category a bit. 
#Investors drawn to Max DTI appear to be drawn to those who are 
#self-employed.  

```

Figure 21.  Investors by employment status where debt to income ratio = 10.01
            and occupation = "Other"
            
Figure 21 expands on Figure 19 & 20 by breaking down the other category
into employment status.  Here we can see investors are drawn to self-employed
borrowers with a high debt to income.  My assumption is that these are 
business loans and investors are taking a chance on a business idea.  

Could homeowner status also have an impact on these high risk loans?  

```{r echo=FALSE, Bivariate_Plots_9, warning=FALSE}

#Plot investors by home ownership status for those with max DTI
p1 <- ggplot(data=subset(loans,loans$DebtToIncomeRatio==10.01 & loans$Occupation=="Other"),
       aes(x=Investors,y=IsBorrowerHomeowner)) +
  geom_col() + 
  ylab("homeowner") + 
  labs(title = "Investors by homeownership for 10.01 debt to income")

p2 <- ggplot(data=loans,
       aes(x=Investors,y=IsBorrowerHomeowner)) +
  geom_col() + 
  ylab("homeowner")+ 
  labs(title = "Investors by homeownership for all loans")

grid.arrange(p1,p2)

########################################################################


```

Figure 22.  Investors by homeowner status for high risk loans and all loans.  

Figure 22 shows that high risk loans given to homeowners have a higher 
proportion of investors than the entire population of loans.  However in 
both cases that Prosper lending system favors homeowners. 

Expanding more on Figure 19. Number of investors by debt to income ratio
The population is so heavy in the 0 - 0.1 range, that zooming in may shed some
light.  

```{r echo=FALSE, Bivariate_Plots_10, warning=FALSE}

#Group by DTI
total_investors_by_DTI <- 
  loans %>%
  group_by(DebtToIncomeRatio) %>%
  summarize(investors = sum(Investors),
            mean= mean(Investors),
            n=n()) %>%
  arrange(DebtToIncomeRatio) %>%
  mutate(id = row_number(),rt=cumsum(n))

#102838 loans have a DTI or 0.6 or less. 
#sqldf ('select count(*) from loans where DebtToIncomeRatio <= 0.60000 ')


#Plot mean investors by DTI demonstrating the density of loans at low DTI
#references:
#https://www.reddit.com/r/rstats/comments/9ecr6g/possible_to_add_vertical_lines_for_1st_and_3rd/
#https://stackoverflow.com/questions/18091721/align-geom-text-to-a-geom-vline-in-ggplot2

#total_investors_by_DTI

ggplot(data=total_investors_by_DTI,
      aes(x=DebtToIncomeRatio)) +
      geom_point(aes(y=investors)) +
      scale_x_continuous(breaks=c(0,0.6,2.5,5.0,7.5,10.01)) +
      scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3)) +
      geom_smooth(aes(y=investors)) +
      ylab("investors") +
      xlab("debt to income ratio") +
      labs(title = "Investors by Debt to Income Ratio") 
      



ggplot(data=subset(total_investors_by_DTI,DebtToIncomeRatio <= 0.6),
      aes(x=DebtToIncomeRatio)) +
      geom_point(aes(y=investors)) +
      scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3)) +
      geom_smooth(aes(y=investors)) +
      ylab("investors") +
      xlab("debt to income ratio") +
      labs(title = "Investors by Debt to Income Ratio") 
      

```


Figure 23.  Debt to income ratio by number of investors
Figure 24.  Debt to income ratio by number of investors for 90% of loans.  

Figure 23 & 24 show that Prosper and investors favor borrowers with a 
debt to income of 0.6 or less.  


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I found that investors weren't drawn to a particular loan status.  

They do seem to be drawn to particular professions, but for the most part
those professions also have high credit scores.  Doctor, Judge, Pharmacist, 
Pilot, Computer Programmer, Professors, Scientist, are examples of professions 
which both have a high number of investors and a high credit score. 
Perhaps more interesting is Fireman which shows on the list of top investors,
but not on the list of top credit scores.  

There does appear to be a correlation between number of investors and
credit score which is not surprising.  Looking at Debt to Income Ratio (DTI), 
Investors have a strong bias for low DTI, which also is not surprising.  
However, I was very surprised to see a spike at the extreme high end of DTI.  
According to the documentation, 10.01 is a cap for DTI and it represents
more than 100% debt to income.  I would assume these borrowers have no chance
of paying back the loan and should not have any investors.  

I zoomed into the data to look at those occupations which have extremely high 
DTI and found the majority to be self-employed.  I believe these are business
loans and the investors must be interested in taking a risk on a business idea. 
I can't tell from the documentation if the investors get to see a business
plan before choosing a loan or if they have to choose based solely on 
borrower's statisticss. 

Finally, I found a large difference between investors choosing homeowners 
over non-homeowners.  One assumption that can be made, at least in the 
United States, is that homeowners tend to be a bit older and more established.  
According to nerdwallet.com, prosper loans are very popular for 
home improvement loans which could be another explanation.  
Perhaps these two factors compound to make such a striking difference.  A
another possibility is that the housing bubble pre-2009 lead to a frenzy of 
home equity loans and investors.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?

Number of total investors by home ownership status. 


# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plot_block_1, warning=FALSE}

###########################################################
#Build a dataset for Multivariate Plot #1 which will compare 
#loans with a DTI ceiling with loans below the ceiling per year
#Was there a feeding frenzy of high risk loans leading up the 2008 recession?
##

#create a grouping of loans per year where the DTI is at the ceiling
mean_investors.year.max.dti <- 
  subset(loans,loans$DebtToIncomeRatio == 10.01) %>%
  group_by(ListingCreationYear) %>%
  summarize(mean_investors = round(mean(Investors),digits=1),
            median_investors = median(Investors),
            n=n()) %>% mutate(event="max") %>%
  arrange(ListingCreationYear)

#create a grouping of loans per year where the DTI is below the ceiling
mean_investors.year.regular.dti <- 
  subset(loans,loans$DebtToIncomeRatio < 10.01) %>%
  group_by(ListingCreationYear) %>%
  summarize(mean_investors = round(mean(Investors),digits=1),
            median_investors = median(Investors),
            n=n())  %>% mutate(event="regular") %>%
  arrange(ListingCreationYear)

#add records for the years where there were no loans above the ceiling
#this keeps the graph easy to read
missing_values <- data.frame(c(2005,2009,2014),
                            c(0,0,0),
                            c(0,0,0),
                            c(0,0,0),
                            c('max','max','max'))
names(missing_values) <- c("ListingCreationYear","mean_investors","median_investors","n","event")

#combine the two previous groupings into one dataframe
mean_investors.year <- 
    rbind(mean_investors.year.regular.dti,mean_investors.year.max.dti,missing_values)

#add to each row total loans for that year for comparison
loans.year <- subset(loans,loans$DebtToIncomeRatio > 0) %>% 
            group_by(ListingCreationYear) %>%
            summarize(total = n())


#join mean_investors.yeat dataset to loans.year dataset.  
#also add spaces to get percent lables to line on either side of the 
#of the y axis
mean_investors.year <-
          sqldf('select y.*,l.total,
                case
                  when event = "regular" then "     "
                END
                || round((y.n/l.total)*100,2) 
                || "%"
                as percent 
                from "mean_investors.year" y inner join 
              "loans.year" l on y.ListingCreationYear = l.ListingCreationYear')


#Completed dataframe for Final Plot #1 
#mean_investors.year

#Plot 1

#mean_investors.year

ggplot(data=mean_investors.year,
      aes(x=factor(ListingCreationYear),y=n,fill=event)) +
      geom_col(position = 'dodge') +
      scale_x_discrete(position = "bottom") + 
      geom_text(aes(label=percent),size = 3,angle=70,vjust=1.5) + 
      xlab("Listing Creation Year") + 
      ylab("Investors") + 
      labs(title ="Investors by Creation Year",
           subtitle = "with percentage of total regular loans")
#this groph shows a yearly trend of mean investors peaking in 2009.
#It shows a surprisingly high investor mean for loans where the DTI as 
#beyond the max, however these represent a tiny percentage of the loans.
###########################################################

```

Figure 25. mean investors per year for normal and high risk loans

This graph shows the number of investors per year.  It is split into 
two parts, those which are regular borrowers (below max debt to income ratio)
and those which are high risk borrowers (all debt and no income). The 
graph also shows in text the percentage of total loans which are to regular borrowers.  

The purpose is to show a feeding frenzy of investors looking for high risk
investments leading up to the recession.  However, this graph doesn't fit 
the narrative.  Loans and Investors rise quickly after the recession 
and the percentage of loans to borrowers with 100% debt to income is so tiny 
that it barely shows on this scale.  

We saw in previous graphs some speculators going after high risk loans 
of self-employed borrowers, but that is clearly a rare occurrence.  


```{r echo=FALSE, Multivariate_Plot_block_2, warning=FALSE}
ggplot(data=subset(loans,
                   DebtToIncomeRatio <= 0.60000 & 
                   Investors <= 750 & 
                   Investors > 0),
      aes(x=DebtToIncomeRatio,y=Investors,group=IsBorrowerHomeowner,color=IsBorrowerHomeowner))+
      geom_point(alpha=1/10) + 
      #scale_y_continuous(labels = label_number(suffix = "", scale = 1e-6)) +
      xlab ("debt to income ratio") +
      ylab ("investors per loan") +
      labs (title = "Investors by debt to income ratio and homeowner status",
            color = "Homeowner") + 
      geom_smooth()

```

Figure 26.  Investors by debt to income ration and homeowner status.  

This graph shows a preference for a borrower with debt to income ratio 
between 1.5 and 2.5 and is a homeowner.  


```{r echo=FALSE, Multivariate_Plot_block_3, warning=FALSE}

#Group by DTI
#85k loans have prosper score
#2009 and later
prosper_scores <- subset(loans,!is.na(ProsperScore) & !is.na(DebtToIncomeRatio))

prosper_score_by_DTI <- 
  prosper_scores %>%
  group_by(ProsperScore) %>%
  summarize(mean= mean(DebtToIncomeRatio),
            median= median(DebtToIncomeRatio),
            n=n()) %>%
  arrange(ProsperScore)

prosper_score_by_credit <- 
  prosper_scores %>%
  group_by(ProsperScore) %>%
  summarize(mean= mean(CreditScoreRangeUpper),
            median= median(CreditScoreRangeUpper),
            n=n()) %>%
  arrange(ProsperScore)

p1<- ggplot(data=prosper_score_by_DTI,
       aes(x=ProsperScore,y=mean)) +
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9')) +
  ylab("mean debt to income ratio") +
  xlab("Prosper score") +
  labs (title = "mean debt to income by Prosper score")
  
p2<- ggplot(data=prosper_score_by_credit,
       aes(x=ProsperScore,y=mean)) +
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9')) +
  scale_y_continuous(limits=c(650,800),oob = rescale_none) +
  ylab("mean credit score") +
  xlab("Prosper score")+
  labs (title = "mean credit score by Prosper score") 
  
  

grid.arrange(p1,p2) 

#completed dataset for final plot #3
#head(prosper_score_by_DTI)

```

Figure 27.  Debt to Income and Credit Score by Prosper score.  

This graph shows that with the introduction of the Prosper score in 2009,
risk assessment continued to use debt to income ratio and credit score
as a factor.  The two variables (credit score and debt to income) follow
nearly identical patterns when compared to Prosper score.  

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Looking at investors by debt to income ratio and investors per year 
investing in high risk loans, I came to the conclusion that Prosper investors
are by and large looking for a safe investment, not wild speculation.  There
are a few exceptions, but those loans are less than 1% of the population.

The introduction of Prosper score in 2009 supports that analysis by 
prosper giving a better score to borrowers with a high credit score and high 
debt to income ratio.

### Were there any interesting or surprising interactions between features?

I was surprised by the consistency for investors to favor homeowners across
the entire range of debt to income.  That can't be by chance. Another 
variable must be influenced by home owner status which is in turn influencing
the number of loans created and the number of investors.  .  

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One, warning=FALSE}
#Plot 1

ggplot(data=mean_investors.year,
      aes(x=factor(ListingCreationYear),y=n,fill=event)) +
      geom_col(position = 'dodge') +
      scale_x_discrete(position = "bottom") + 
      geom_text(aes(label=percent),size = 3,angle=70,vjust=1.5) + 
      xlab("Listing Creation Year") + 
      ylab("Investors") + 
      labs(title ="Investors by Creation Year",
           subtitle = "with percentage of total regular loans")
#

```

### Description One

I chose the breakdown of investors per year split into regular and maximum
debt to income ratio.  We showed earlier how a few high risk loans can 
attract a massive number of investors, however, this is clearly not the norm.  
It would be interesting to draw this same graph with data leading up to present 
day and see if the trend continues.  It would also be interesting to have 
rolling datato see when investors get on board with a loan.  When a loan reaches 
some critical mass of investors, do others simply follow the herd?  

### Plot Two

```{r echo=FALSE, Plot_Two, warning=FALSE}

ggplot(data=subset(loans,
                   DebtToIncomeRatio <= 0.60000 & 
                   Investors <= 750 & 
                   Investors > 0),
      aes(x=DebtToIncomeRatio,y=Investors,group=IsBorrowerHomeowner,color=IsBorrowerHomeowner))+
      geom_point(alpha=1/10) + 
      #scale_y_continuous(labels = label_number(suffix = "", scale = 1e-6)) +
      xlab ("debt to income ratio") +
      ylab ("investors per loan") +
      labs (title = "Investors by debt to income ratio and homeowner status",
            color = "Homeowner") + 
      geom_smooth()


```

### Description Two

For plot 2, I chose to show debt to income ratio vs investors
for borrowers with a DTI of 0.6 or less, separated by homeowner status.  
I chose 0.6 because 90% of all loans fall within this range.  This graph 
shows that both prosper and investors are attracted to borrowers with a 
certain range of debt to income between 0.15 and 0.2.  This shows investors
are looking for a safe investment.  

We don't have data for individual investors. Therefore, if a single investor
invests in multiple loans, he or she is counted many times.  It would be 
interesting to know about the investor.  Could we then predict which 
investors go after which type of loans?  Could we gain insights into how 
to market to low and high risk investors differently to attract more? 

### Plot Three

```{r echo=FALSE, Plot_Three, warning=FALSE}

p1<- ggplot(data=prosper_score_by_DTI,
       aes(x=ProsperScore,y=mean)) +
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9')) +
  ylab("mean debt to income ratio") +
  xlab("Prosper score") +
  labs (title = "mean debt to income by Prosper score")
  
p2<- ggplot(data=prosper_score_by_credit,
       aes(x=ProsperScore,y=mean)) +
  geom_bar(stat='identity',color = I('black'), fill = I('#099DD9')) +
  scale_y_continuous(limits=c(650,800),oob = rescale_none) +
  ylab("mean credit score") +
  xlab("Prosper score")+
  labs (title = "mean credit score by Prosper score") 
  
  

grid.arrange(p1,p2) 

```

### Description Three

Finally, we see that the prosper score is related to both 
debt to income ratio and credit score.  The higher the 
prosper score, the lower the risk.  Prosper is using this score to advise 
investors.  Therefore, going forward with more recent data, we should be 
able to predict the number of investors based on the prosper score more so 
than other factors such as occupation.  

------

# Reflection

Investors are looking for a safe return on investment with the exception 
of a very small number of high risk loans.  
We don't know the amount that each investor put toward each loan
and we don't know the number of distinct investors.  Prosper score was added
after the 2008 recession for the purpose of guiding investors and is using
at least some of the same variables which investors were using individually 
prior to 2009.  

Future work could look into whether or not investors are choosing wisely.  
The fields preceeded with "LP_" 
- LP_CustomerPayments
- LP_CustomerPrincipalPayments
- LP_InterestandFees
- LP_ServiceFees
- LP_CollectionFees
- LP_GrossPrincipalLoss
- LP_NetPrincipalLoss
- LP_NonPrincipalRecoverypayments
Could be analyzed to determine if investors are getting a strong return and 
which investors experienced opportunity cost by investing in the wrong loans. 
Compared the market in general, is Prosper a good choice for their portfolio?


With more information about the small number of high risk loans and the 
investors attracted to them, we could determine which high risk loans, if any,
future investors should consider. 

# References:

https://www.nerdwallet.com/reviews/loans/personal-loans/prosper-personal-loans

https://discuss.analyticsvidhya.com/t/how-to-count-number-of-distinct-values-in-a-column-of-a-data-table-in-r/1124/2

https://stackoverflow.com/questions/50935857/ggplot2-keeps-adding-5-at-end-of-year-variable-on-x-axis

https://www.roelpeters.be/scale-ggplot-y-axis-millions-or-thousands-r/
